Building socket-aware BPF programs

### 2018-10-27 paper/paper.tex
paper: More fleshing out

@@

This calls for the BPF infrastructure to handle reference tracking, and to hide
this detail from the BPF API. Whenever a socket lookup function is called, a
reference is taken and the socket is added to a reference list. At the end of
the BPF program execution, the core code could walk this reference list and
release each reference. However, collecting references and releasing them at
the end of the BPF program invocation has the unfortunate overhead that even
the execution of BPF programs that do not make use of the new socket lookup
helpers would need to pay the cost of checking the list of references to free.
When dealing with sufficiently demanding use cases such as those which use XDP,
even the few instructions required to implement such a check may have a
measurable impact on performance.

@@

Some instruction calls are restricted while holding a resource reference to
avoid leaking references, for example the \verb+bpf_tail_call+ helper function.

@@

The verifier is tasked with ensuring that all BPF program execution flow is
safe to the extents where it interacts with the BPF API. In general, BPF
programs, once verified and triggered by event hooks, will execute from the
first instruction through the final instruction for a particular path through
the program. Therefore, after execution of a BPF program containing resource
acquire and release functions which are verified by the logic above, the
underlying resources should not leak.

One exception to this is when classic BPF \verb+LD_ABS+ or \verb+LD_IND+
instructions are used in the BPF program. These instructions were previously
used to provide direct packet access, with the semantics that if the access
offset exceeds the length of the packet, the BPF program would be terminated
prior to the final state. Without mitigation, this could lead to leaking of
references. Newer BPF APIs provide better alternatives for packet access, so we
disallow the use of these instructions while holding a socket reference.

@@

To implement explicit reference tracking, the API required at least two
functions to handle referencing of sockets from the BPF API: A socket lookup
function and a socket release function. This section describes the API
considerations for each of these.

@@

Linux network devices are associated with a particular network namespace which
provides logical network stack separation. Sockets are also inherently
associated with an application that operates in a particular network namespace.
The simplest path for handling namespacing from the lookup handlers would be to
allow BPF programs operating on a device within a network namespace to only
find sockets in the same network namespace. By passing the BPF program context
into the socket lookup helper, the implementation can derive the source network
namespace from this context. Beyond this, with the growing popularity of
containers it may be desirable for a network orchestration tool to make use of
socket properties inside container network namespaces to influence decisions
made on packets outside the container---for instance, on packets being received
into a node prior to the packet passing into the container network namespace,
or if filtering is implemented on a veth pair outside of a container. By
allowing the netns ID to be specified, this use case can be supported as well.

@@

A common use case for the socket lookup functions would call them from a packet
context, where the \verb+__sk_buff+ has the full 5-tuple information available
and easily accessible. One could imagine a simpler helper function that allows
the BPF program writer to provide the \verb+__sk_buff+ to the helper function,
then the helper function would look up the socket based upon the packet
metadata. This would however limit the potential uses for such a helper. Two
cases that would be more limited with this model are usage from XDP, or if the
BPF program implements a form of network address translation. From XDP, if the
raw \verb+xdp_md+ context is provided to such a helper, then the helper would
need to parse the packet to understand the 5-tuple. This would duplicate the
standard stack paths which are executed after XDP where the \verb+sk_buff+ is
built, but this information would be subsequently thrown away upon return of
the helper call. Adding to this, if the program performs network address
translation between the local application and the remote destination, then the
local stack may not contain a socket associated with the tuple that is directly
in the packet. While these cases could be worked around with a simpler helper,
it was deemed more powerful and generic to allow the BPF program writer to
provide the tuple for lookup.

@@

New kernel APIs that have any scope for extension should contain a flags
argument. For the socket lookup helpers, a few ideas had been considered as
potential future alternatives to the existing behaviour. The lookup functions
follow the standard socket lookup paths in Linux which have predetermined
methods for selecting a socket when the application uses \verb+SO_REUSEPORT+.
Some discussion on the mailinglist proposed allowing BPF program writers to
influence the socket selection mechanism~\cite{netdev-sk-select}.

@@

Another aspect of extensibility is the ability to lookup sockets which are not
TCP or UDP. The initial RFC of this patch series proposed a single lookup
function which would choose the L4 protocol based upon a field in the
tuple~\cite{sk-lookup-rfc}, however this would make it more difficult for BPF
program writers to detect the support for different protocol types at compile
time. There is not a significant number of supported L4 socket types in the
Linux stack today, and the number is not expected to increase drastically, so
it was considered simpler and easier to split out each L4 socket lookup
function into an independent helper call.

@@

For some socket types, such as UDP sockets, or TCP listen sockets, memory
access safety can be achieved with minimal implementation: The destruction of
such sockets is already governed by standard RCU rules, meaning that while the
RCU lock is held, they can be safely referenced; Once the RCU grace period is
reached, the memory may be freed and references to the socket are no longer
safe to use. Because BPF programs run under the RCU lock, this allows simple,
safe access of properties of these sockets without requiring additional
mechanisms to ensure the safety of the socket pointers.

When it comes to TCP sockets that are not governed by RCU, the networking stack
uses reference counting to manage socket memory instead. To ensure that the
socket remains safe to access for a period of BPF program execution, the BPF
program must also take a reference on the socket. This also implies that such
references must be subsequently released to prevent leaking kernel resources.

@@

The tuple structure is defined in such a way that if the IPv4 or IPv6 packet is
immediately followed by the TCP or UDP header without IP options in between,
then a pointer directly to the L3 addresses may be passed to the lookup
function, allowing the helper function to directly pull the addresses and ports
from the packet buffer rather than requiring the BPF program to first extract
these onto the stack.

@@

Stateful firewalling is distinguished from stateless firewalling in that it
provides the ability to associate two directions of a connection together and
apply a filtering policy based upon the direction of the connection. This
commonly involves using a connection tracker to reconstruct the connection
state of each endpoint of the connection without specific knowledge from those
endpoints~\cite{ovs-ct,conntrack}. In cases where the Linux stack acts as a
firewall on a node in the network that is separate from both endpoints, In many
cases this model is used when filtering traffic where one endpoint is colocated
with the Linux stack that performs the firewalling, which already contains
information about the local end of the connection in a slightly.

@@

The core BPF maintainers provided great guidance and review of the patches, in
particular Alexei Starovoitov, Daniel Borkmann and Martin KaFai Lau.

>>>>

%To ensure that the socket remains safe to access for a period of BPF program
%execution, the BPF program must also take a reference on the socket. This also
%implies that such references must be subsequently released to prevent leaking
%kernel resources.

@@

This calls for the BPF infrastructure to handle reference tracking, and to hide
this detail from the BPF API. Whenever a socket lookup function is called, a
reference is taken and the socket is added to a reference list. At the end of
the BPF program execution, the core code could walk this reference list and
release each reference. However, collecting references and releasing them at
the end of the BPF program invocation has the unfortunate overhead that even
the execution of BPF programs that do not make use of the new socket lookup
helpers would need to pay the cost of checking the list of references to free.
When dealing with sufficiently demanding use cases such as those which use
eXpress Data Path (XDP)~\cite{xdp}, even the few instructions required to
implement such a check may have a measurable impact on performance.

@@

Some instruction calls are restricted while holding a resource reference to
avoid leaking references, for example the \verb+bpf_tail_call+ helper function,
which would otherwise leak the pointer reference to a subsequent BPF program.

@@

The verifier is tasked with ensuring that all BPF program execution flow is
safe to the extents where it interacts with the BPF API. In general, BPF
programs, once verified and triggered by event hooks, will execute from the
first instruction through the final instruction for a particular path through
the program. Therefore, after execution of a BPF program containing resource
acquire and release functions which are verified by the logic above, references
to those resources should not leak.

One exception to this is when classic BPF \verb+LD_ABS+ or \verb+LD_IND+
instructions are used in the BPF program. These instructions were previously
used to provide direct packet access, with the semantics that if the access
offset exceeds the length of the packet, the BPF program would be terminated
prior to the final state. Without mitigation, this could lead to leaking of
references at runtime, even if the instructions appear to balance acquisition
and release at verification time. Newer BPF APIs provide better alternatives to
these instructions for packet access, so we disallow the use of these
instructions while holding a socket reference.

@@

To provide an API that implements explicit reference tracking, at least two
functions are needed to handle referencing of sockets: A socket lookup function
and a socket release function. This section describes the API considerations
for each of these.

@@

Linux network devices are associated with a particular network namespace which
provides logical network stack separation~\cite{netns, netns-netdevice}.
Sockets are also inherently associated with an application that operates in a
particular network namespace~\cite{netns-sock}. The simplest path for handling
namespacing from socket lookup handlers operating from a network device would
be to allow BPF programs operating on a device within a network namespace to
only find sockets in the same network namespace. By passing the BPF program
context into the socket lookup helper, the implementation can derive the source
network namespace from this context. Beyond this, with the growing popularity
of containers it may be desirable for a network orchestration tool to make use
of socket properties inside container network namespaces to influence decisions
made on packets outside the container---for instance, on packets being received
into a node prior to passing the packet into the network namespace. By allowing
the netns ID to be specified, this use case can be supported as well.

@@

A common use case for the socket lookup functions would call them from a packet
context, where the \verb+__sk_buff+ has the full 5-tuple information available
and easily accessible. One could imagine a simpler helper function that allows
the BPF program writer to provide the \verb+__sk_buff+ to the helper function,
then the helper function would look up the socket based upon the packet
metadata. This would however limit the potential uses for such a helper. Two
cases that would be more limited with this model are usage from XDP (which does
not inherently parse packets or collect packet metadata into the context), or
if the BPF program implements a form of network address translation. From XDP,
if the raw \verb+xdp_md+ context is provided to such a helper, then the helper
would need to parse the packet to understand the 5-tuple. This would duplicate
the standard stack paths which are executed after XDP where the \verb+sk_buff+
is built, but this information would be subsequently thrown away upon return of
the helper call. Adding to this, if the program performs network address
translation between the local application and the remote destination, then the
local stack may not contain a socket associated with the tuple that is directly
in the packet. While these cases could be worked around with a simpler helper,
it was deemed more powerful and generic to allow the BPF program writer to
provide the tuple for lookup.

@@

New kernel APIs that have any scope for extension should contain a flags
argument. For the socket lookup helpers, a few ideas had been considered as
potential future alternatives to the existing behaviour. The lookup functions
follow the standard socket lookup paths in Linux which have predetermined
methods for selecting a socket when the application uses \verb+SO_REUSEPORT+.
Some subsequent discussion on the mailinglist proposed allowing BPF program
writers to influence the socket selection mechanism~\cite{netdev-sk-select}.

@@

Another aspect of extensibility is the ability to lookup sockets which are not
TCP or UDP. The initial RFC of this patch series proposed a single lookup
function which would choose the Layer~4 protocol based upon a field in the
tuple~\cite{sk-lookup-rfc}, however this would make it more difficult for BPF
program writers to detect the support for different protocol types at compile
time. When a socket type is unsupported, the implementation would return
\verb+NULL+, implying that there is no such socket. There is not a significant
number of supported Layer~4 socket types in the Linux stack today, and the
number is not expected to increase drastically, so it was considered simpler
and easier to split out each Layer~4 socket lookup function into an independent
helper call, so that the result could simply return a socket or \verb+NULL+ and
not need to encode a representation of other error conditions.

@@

For some socket types, such as UDP sockets, or TCP listen sockets, memory
access safety can be achieved with minimal implementation: The destruction of
such sockets is already governed by standard RCU rules, meaning that while the
RCU lock is held, they can be safely accessed without holding a reference; Once
the RCU grace period is reached, the memory may be freed and references to the
socket are no longer safe to use. For these socket types, since BPF programs
run under the RCU lock, the properties of these sockets can be accessed
directly without taking a reference on the socket. As such, the implementation
of the socket lookup and release functions can avoid the atomic reference count
increment operation.

The API retains explicit reference tracking to ensure consistency of the API
and to allow multiple underlying implementations to handle reference tracking
in a way that is safe for the implementation. For instance, for  TCP sockets
that are not governed by RCU, the networking stack uses reference counting to
manage socket memory instead, so the socket lookup and release functions would
take and release references on the socket, respectively.

@@

The tuple structure is defined in such a way that if the IPv4 or IPv6 packet is
immediately followed by the TCP or UDP header without IP options in between,
then a pointer to the packet data at the offset of the Layer~3 addresses may be
passed to the lookup function, allowing the implementation to directly pull the
addresses and ports from the packet buffer rather than requiring the BPF
program to first extract these onto the stack and pass a pointer to the stack
copy of the tuple.

@@

Stateful firewalling is distinguished from stateless firewalling in that it
provides the ability to associate two directions of a connection together and
apply a filtering policy based upon the direction of the connection. This
commonly involves using a connection tracker to reconstruct the connection
state of each endpoint of the connection without specific knowledge from those
endpoints~\cite{conntrack,ovs-ct}. In cases where Linux is configured as a
firewall between applications that exist in distinct Linux kernel copies
running in the network, this makes sense; The information that is needed to
perform stateful firewalling is not present on the Linux instance that needs
this information, so the information must either be shared from the peers
(assuming that the firewall orchestrator has control over the peer endpoints),
or this information must be synthesised based upon the packets that are seen on
this intermediate Linux instance. However, in many cases this model is used
when filtering traffic where one endpoint is co-located with the Linux stack
that performs the firewalling, which already contains information about the
local end of the connection. In this use case, the connection tracking table
effectively duplicates knowledge that Linux already creates and maintains.

Using the socket lookup functions described in this paper, in conjunction with
the socket properties including local Layer~3 and Layer~4 information available
in the \verb+bpf_sock+ (Listing~\ref{lst:helper-struct}), BPF programs can
identify the directionality of connections for packets ingressing or egressing
the Linux stack on a network device.

@@

The core BPF maintainers provided great guidance and review of the patches, in
particular Alexei Starovoitov, Daniel Borkmann and Martin KaFai Lau. Additional
shout-out to Nitin Hande who provided testing and feedback.
